---
title: "Module 1: Data Setup and Inspection"
subtitle: "CRP 241 — IVAM-ED Teaching Case"
author: "Steve Grambow, PhD"
date: today
execute:
  echo: true       # show all code in the rendered document
  warning: false
  message: false
---

## Overview

This document introduces the IVAM-ED synthetic dataset and walks through the first
analytical step in any research project: **understanding your data before analyzing it**.

By the end of this module you should be able to:

1. Load a dataset in R and inspect its structure.
2. Identify variable types (continuous, binary, ordinal) and their scales.
3. Count missing values and understand why they occur.
4. Recognize the difference between baseline measurements and follow-up measurements.

---

## The IVAM-ED Study

The **IVAM-ED** trial evaluated whether a voice-activated smart speaker (Amazon Echo
Dot) could improve mental health and diabetes self-management among older adults with
type 2 diabetes. Researchers in Brazil randomly assigned 112 adults (age ≥ 65) to
either:

- **Intervention:** Use a smart speaker at home for 12 weeks. The device was
  programmed with reminders, health messages, and educational podcasts.
- **Usual care:** Continue with standard diabetes management and no device.

The **primary outcome** was the SRQ-20 score (Self-Reporting Questionnaire, 0–20),
a measure of mental distress. Lower scores mean less distress.

> **Important:** We are working with a **synthetic (simulated) dataset** — not real
> participant data. The dataset was constructed to approximate the published summary
> statistics. See [simulation-assumptions.md](../simulation-assumptions.md) for how
> it was built and what assumptions were made.

---

## Setup

```{r}
#| label: setup

# Load packages
# If any of these are missing, run: renv::install("package_name")
library(dplyr)    # data manipulation
library(labelled) # reading variable labels
```

---

## 1. Generate the Synthetic Dataset

The simulation script creates the dataset from scratch using a fixed random seed,
so you will always get the same data.

```{r}
#| label: generate-data

# Source the simulation script. This creates data/ivam_synthetic.rds.
# The verification summary will print below.
source("simulate-ivam-ed.R")
```

::: {.callout-note}
**Why the data are always the same: `set.seed()`**

At the top of `simulate-ivam-ed.R` you will find `set.seed(241)`. This locks
R's random number generator to a fixed starting point. R's "randomness" is
actually a deterministic algorithm — a mathematical formula that produces a long
sequence of numbers that *look* random but are completely determined by the
starting value (the seed). Given the same seed, R will always generate the same
sequence.

This is deliberate for a teaching dataset: every student gets identical data,
so results are comparable and answers are verifiable.

**Try it yourself:** Open `R/simulate-ivam-ed.R`, change `set.seed(241)` to
`set.seed(999)`, source the script, and re-run this module. Your Table 1 means
will shift slightly, the primary ANCOVA result may change in magnitude, and it
may or may not remain statistically significant — illustrating that any single
study is one realization of a random process.
:::


---

## 2. Load and Inspect the Data

```{r}
#| label: load-data

# Load the saved dataset
dat <- readRDS("../data/ivam_synthetic.rds")
```

### 2a. Dimensions

```{r}
#| label: dimensions

# How many rows (participants) and columns (variables)?
dim(dat)
```

::: {.callout-note}
**What to notice:** Each row is one participant. Each column is one variable. The
number of rows should equal the number of enrolled participants (112).
:::

### 2b. Variable list with labels

The dataset includes human-readable labels on every variable. Use `look_for()` to
see them all at once:

```{r}
#| label: variable-list

# Show all variable names and their labels
labelled::look_for(dat)
```

::: {.callout-tip}
**Reading the label output:** The `label` column tells you what each variable means,
including the score direction (higher = better vs. higher = worse) and range.
This is important — not all instruments point the same way!
:::

### 2c. Data types

```{r}
#| label: data-types

# str() shows the type (numeric, integer, character) and first few values
str(dat)
```

::: {.callout-note}
**What to notice:** Variables like `group`, `female`, `insulin`, `hx_depression`,
and `hx_anxiety` are stored as integers (0 or 1). They represent binary categories,
not truly numeric quantities. We'll convert these to labeled factors in later modules.
:::

---

## 3. Understand the Randomization

```{r}
#| label: group-counts

# Count participants per group
# group = 1 is intervention (smart speaker); group = 0 is usual care
table(dat$group)
```

```{r}
#| label: group-labels

# A clearer version using dplyr
dat |>
  count(group) |>
  mutate(group_label = ifelse(group == 1, "Smart speaker", "Usual care"))
```

::: {.callout-note}
**What to notice:** The trial used a 1:1 randomization ratio, so both groups
should have exactly 56 participants at enrollment. Simple random assignment
(as used here) does not guarantee perfect balance on every baseline characteristic —
we'll examine that in Module 2.
:::

---

## 4. Identify the Outcome Measures

The dataset contains measurements at two time points:

| Suffix | Time point | Example |
|--------|------------|---------|
| `_0`   | Baseline (week −1, before intervention) | `srq20_0` |
| `_12`  | Follow-up (week 12, end of intervention) | `srq20_12` |

```{r}
#| label: outcome-names

# Select just the outcome columns to see their structure
dat |>
  select(participant_id, group, srq20_0, srq20_12,
         sf36_0, sf36_12, hba1c_0, hba1c_12) |>
  head(10)
```

::: {.callout-tip}
**Score directions to keep in mind:**

- **SRQ-20** (0–20): Higher score = MORE mental distress (bad)
- **SF-36** (0–100): Higher score = BETTER quality of life (good)
- **HbA1c** (%): Higher = WORSE glycemic control (bad)
- **SCI-R** (0–56): Higher = BETTER self-care adherence (good)
- **PSS** (0–40): Higher = MORE perceived stress (bad)
:::

---

## 5. Missing Data

Not everyone completed the 12-week follow-up. In the real trial, 9 participants
(8%) did not return for the follow-up visit — 3 died and 6 were lost to follow-up.

```{r}
#| label: missing-overview

# How many participants are missing at follow-up?
dat |>
  count(missing_12wk, reason_missing)
```

```{r}
#| label: missing-by-group

# Is the missingness distributed equally between the two groups?
dat |>
  group_by(group) |>
  summarise(
    n_enrolled  = n(),
    n_missing   = sum(missing_12wk),
    pct_missing = round(mean(missing_12wk) * 100, 1)
  )
```

```{r}
#| label: missing-outcomes

# Confirm: follow-up outcomes are NA for those with missing_12wk = 1
dat |>
  filter(missing_12wk == 1) |>
  select(participant_id, group, reason_missing, srq20_12, hba1c_12)
```

::: {.callout-warning}
**Why this matters:** Missing data is a recurring challenge in clinical trials.
When participants drop out, we must decide how to handle the missing values in our
analysis. We will explore this in Module 6 (Missing Data). For now, we will
flag that missingness is present and roughly balanced between groups.
:::

---

## 6. Quick Sanity Checks

Before doing any analysis, it is good practice to check for values that fall
outside the expected instrument ranges. These could indicate data entry errors
(in real data) or simulation bugs (in synthetic data).

```{r}
#| label: range-checks

# Check that all scores fall within their valid ranges
# Any TRUE in the "out_of_range" column would indicate a problem

range_checks <- tibble(
  variable  = c("srq20_0",  "srq20_12", "sf36_0",  "sf36_12",
                "sci_r_0", "sci_r_12", "pss_0",   "pss_12",
                "hba1c_0", "hba1c_12", "mmse_0"),
  min_valid = c(0,  0,  0,   0,  0, 0, 0,  0,  4.5, 4.5, 0),
  max_valid = c(20, 20, 100, 100, 56, 56, 40, 40, 18,  18,  30)
)

range_checks |>
  rowwise() |>
  mutate(
    obs_min = min(dat[[variable]], na.rm = TRUE),
    obs_max = max(dat[[variable]], na.rm = TRUE),
    out_of_range = obs_min < min_valid | obs_max > max_valid
  )
```

---

## 7. Save a Clean Working Copy

It is good practice to keep the raw data unchanged and save a separate working
copy with any modifications (like renaming variables or recoding groups).

```{r}
#| label: save-working-copy

# Create a working copy with a descriptive group label
dat_work <- dat |>
  mutate(
    group_label = factor(group,
                         levels = c(0, 1),
                         labels = c("Usual care", "Smart speaker"))
  )

# Save the working copy
saveRDS(dat_work, file = "../data/ivam_working.rds")

cat("Working dataset saved to data/ivam_working.rds\n")
cat("Dimensions:", nrow(dat_work), "rows x", ncol(dat_work), "columns\n")
```

---

## Summary

In this module you:

- Loaded the IVAM-ED synthetic dataset (112 participants, 25 variables)
- Identified the two time points (baseline `_0` and follow-up `_12`)
- Confirmed randomization produced two groups of 56
- Found 9 participants (8%) with missing follow-up data, slightly more in the
  control group (5 vs. 4)
- Verified all score ranges are within valid instrument bounds

**Next:** Module 2 will examine baseline characteristics using descriptive
statistics and explore whether randomization produced balanced groups.

---

## Session Information

```{r}
#| label: session-info

# Record the R version and package versions for reproducibility
sessionInfo()
```
