---
title: "Module 2: Descriptive Statistics and Baseline Characteristics"
subtitle: "CRP 241 — IVAM-ED Teaching Case"
author: "Steve Grambow, PhD"
date: today
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This module covers the first analytical step after loading your data: **describing
the sample**. In a randomized trial, descriptive statistics serve two purposes:

1. **Characterize who was in the study** — so readers know whether the findings
   might apply to other populations.
2. **Check that randomization worked** — by comparing baseline characteristics
   between the intervention and control groups.

By the end of this module you should be able to:

1. Compute means, standard deviations, medians, and frequencies for continuous
   and categorical variables.
2. Construct a formatted baseline characteristics table (Table 1).
3. Explain why we do not perform significance tests of baseline differences in
   a randomized trial.
4. Distinguish between clinically meaningful and trivially small group differences
   at baseline.

---

## Setup

```{r}
#| label: setup

library(dplyr)
library(gtsummary)  # produces publication-ready Table 1
library(labelled)   # variable labels
```

```{r}
#| label: load-data

# Load the working copy created in Module 1
# If data/ivam_working.rds does not exist, run R/simulate-ivam-ed.R first,
# then render 01-data-setup.qmd to generate the working copy.
dat <- readRDS("../data/ivam_working.rds")
```

---

## 1. Univariate Summaries

Before building a grouped table, it helps to look at the whole sample
together. This gives you a feel for the data and catches obvious problems
(values out of range, unexpected distributions).

### Continuous variables

```{r}
#| label: continuous-summaries

# Summarize all continuous outcome and clinical variables
dat |>
  select(age, edu_years, income_wages, dm_duration_y,
         mmse_0, srq20_0, sci_r_0, pss_0, sf36_0, hba1c_0) |>
  summarise(across(everything(),
    list(
      n     = \(x) sum(!is.na(x)),
      mean  = \(x) round(mean(x, na.rm = TRUE), 1),
      sd    = \(x) round(sd(x,   na.rm = TRUE), 1),
      min   = \(x) round(min(x,  na.rm = TRUE), 1),
      max   = \(x) round(max(x,  na.rm = TRUE), 1)
    ),
    .names = "{.col}__{.fn}"
  )) |>
  tidyr::pivot_longer(everything(),
    names_to  = c("variable", "stat"),
    names_sep = "__"
  ) |>
  tidyr::pivot_wider(names_from = stat, values_from = value)
```

::: {.callout-tip}
**What `pivot_longer()` and `pivot_wider()` do:** R stores data in either *wide*
or *long* format. Wide format has one row per participant with many columns.
Long format has one row per observation, with a column identifying what variable
that row describes.

We used `pivot_longer()` above to convert the `summarise()` output — which has
one column per statistic — into a long table where each row is one
(variable, statistic) pair. Then `pivot_wider()` re-shapes it so each statistic
becomes its own column. This pattern appears throughout the modules whenever we
need to reshape data for a cleaner table or a grouped plot.
:::

::: {.callout-note}
**What to notice:** Compare the means and SDs to the published Table 1 in
Matzenbacher et al. (2026). The pooled (overall) values should be close.
Small differences are expected because this is a simulation.
:::

### Categorical variables

```{r}
#| label: categorical-summaries

# Count and percentage for each binary variable
binary_vars <- c("female", "insulin", "hx_depression", "hx_anxiety",
                 "cognitive_decline")

dat |>
  select(all_of(binary_vars)) |>
  summarise(across(everything(),
    list(
      n   = \(x) sum(x == 1, na.rm = TRUE),
      pct = \(x) round(mean(x == 1, na.rm = TRUE) * 100, 1)
    ),
    .names = "{.col}__{.fn}"
  )) |>
  tidyr::pivot_longer(everything(),
    names_to  = c("variable", "stat"),
    names_sep = "__"
  ) |>
  tidyr::pivot_wider(names_from = stat, values_from = value) |>
  mutate(summary = paste0(n, " (", pct, "%)"))
```

---

## 2. Table 1: Baseline Characteristics by Randomized Group

The standard way to present baseline characteristics in a clinical trial is a
**Table 1**: one column per group, with each row showing a baseline variable.

We use `gtsummary::tbl_summary()` to build this table. The package automatically
selects appropriate statistics (mean ± SD for continuous variables, n (%) for
categorical variables) and formats the output for publication.

```{r}
#| label: table1

# Prepare a clean analysis dataset for Table 1
# Rename group to a labeled factor so the column headers read clearly
dat_t1 <- dat |>
  select(
    group_label,
    # Demographics
    age, female, edu_years, income_wages,
    # Clinical
    dm_duration_y, insulin,
    # Comorbidities
    hx_depression, hx_anxiety, cognitive_decline,
    # Baseline scores
    mmse_0, srq20_0, sci_r_0, pss_0, sf36_0, hba1c_0
  ) |>
  mutate(
    # Convert binary integers to labeled factors for cleaner table display
    female            = factor(female,            levels = c(1, 0), labels = c("Female", "Male")),
    insulin           = factor(insulin,           levels = c(1, 0), labels = c("Yes", "No")),
    hx_depression     = factor(hx_depression,     levels = c(1, 0), labels = c("Yes", "No")),
    hx_anxiety        = factor(hx_anxiety,        levels = c(1, 0), labels = c("Yes", "No")),
    cognitive_decline = factor(cognitive_decline, levels = c(1, 0), labels = c("Yes", "No"))
  )

# Build the table
tbl_summary(
  dat_t1,
  by = group_label,
  label = list(
    age               ~ "Age, years",
    female            ~ "Female sex",
    edu_years         ~ "Education, years",
    income_wages      ~ "Income, min. wages/month",
    dm_duration_y     ~ "Diabetes duration, years",
    insulin           ~ "Insulin use",
    hx_depression     ~ "History of depression",
    hx_anxiety        ~ "History of anxiety",
    cognitive_decline ~ "Cognitive decline screening (positive)",
    mmse_0            ~ "MMSE score (0–30)",
    srq20_0           ~ "SRQ-20 score (0–20)",
    sci_r_0           ~ "SCI-R score (0–56)",
    pss_0             ~ "PSS score (0–40)",
    sf36_0            ~ "SF-36 score (0–100)",
    hba1c_0           ~ "HbA1c, %"
  ),
  statistic = list(
    all_continuous()  ~ "{mean} ({sd})",
    all_categorical() ~ "{n} ({p}%)"
  ),
  digits = list(
    all_continuous() ~ 1
  ),
  missing = "no"   # missing data handled separately
) |>
  add_overall() |>
  modify_header(label ~ "**Characteristic**") |>
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Randomized group**") |>
  bold_labels()
```

::: {.callout-note}
**Compare to the published table:** Pull up Table 1 in Matzenbacher et al. (2026).
Your simulated table should have similar overall means and SDs but will not match
exactly — this is expected with a synthetic dataset. The group-level differences
(imbalances) in your table will also differ from the published table because they
arise from random chance in the simulation.
:::

---

## 3. Interpreting Baseline Balance

### 3a. What "balanced" means in a randomized trial

After randomization, we compare baseline characteristics between groups. In a
*perfectly* balanced trial, every characteristic would be identical across groups.
In practice, small differences always occur by chance, even with proper
randomization.

The key question is: **are any of the differences large enough to matter?**

```{r}
#| label: baseline-differences

# Calculate the mean difference between groups for each continuous outcome
# and the standardized mean difference (SMD)
dat |>
  group_by(group_label) |>
  summarise(
    age_mean       = mean(age),
    srq20_mean     = mean(srq20_0),
    hba1c_mean     = mean(hba1c_0),
    sf36_mean      = mean(sf36_0),
    pss_mean       = mean(pss_0),
    sci_r_mean     = mean(sci_r_0)
  ) |>
  tidyr::pivot_longer(-group_label, names_to = "variable", values_to = "mean") |>
  tidyr::pivot_wider(names_from = group_label, values_from = mean) |>
  mutate(difference = round(`Smart speaker` - `Usual care`, 2))
```

### 3b. Why we do NOT test for statistical significance at baseline

You may notice that Table 1 in our simulated output (and in most published trials)
does not include a p-value column. This is intentional, and it is one of the most
important conventions in clinical trial reporting.

**The reason:** In a properly randomized trial, any baseline difference between
groups arose by chance — by definition, because we randomized. Testing whether
a baseline difference is "statistically significant" asks: "Is this difference
too large to be due to chance?" But we already know it is due to chance (the
randomization). The test cannot tell us anything informative.

What matters instead is whether the difference is **clinically meaningful** — large
enough to affect the outcome, regardless of its p-value. A large p-value does not
mean a baseline imbalance is unimportant; it means only that the sample was too
small to detect it with confidence.

::: {.callout-important}
**CONSORT reporting guideline:** Table 1 in a randomized trial should show
baseline characteristics but **should not include p-values for between-group
comparisons**. This is stated explicitly in the CONSORT 2010 guidelines
(Schulz et al., 2010, *BMJ*).
:::

### 3c. Standardized mean differences

A more informative measure of baseline balance is the **standardized mean
difference (SMD)**: the group difference divided by the pooled standard deviation.
Values below 0.1 are generally considered small (good balance); values above 0.2
may warrant adjustment in the primary analysis.

```{r}
#| label: smd

# gtsummary can add SMDs directly to the Table 1
tbl_summary(
  dat_t1 |> select(group_label, age, srq20_0, hba1c_0, sf36_0, pss_0, sci_r_0),
  by = group_label,
  statistic = list(all_continuous() ~ "{mean} ({sd})"),
  digits    = list(all_continuous() ~ 1),
  missing   = "no"
) |>
  add_difference(test = list(all_continuous() ~ "smd")) |>
  modify_header(estimate ~ "**SMD**") |>
  bold_labels()
```

::: {.callout-note}
**Interpreting SMDs in this dataset:** Because we simulated from pooled group
means (the same mean for both groups), any SMDs you see reflect random variation
from the 1:1 assignment with this particular seed. In the real trial, Table 1
also showed no important baseline differences. The ANCOVA model in Module 5 will
adjust for the baseline value of each outcome regardless of the SMD, providing a
more precise estimate of the treatment effect.
:::

---

## 4. Practice Exercises

1. **Recalculate by hand.** Compute the mean SRQ-20 score for each group using
   only base R (no packages). Compare to the gtsummary table.

2. **The education variable.** In the real trial, 24.1% of participants had
   completed secondary education or higher. What percentage do you find in the
   simulated data? Is it close? (Check
   [simulation-assumptions.md](../simulation-assumptions.md) to see where this
   parameter came from.)

3. **Identify the most imbalanced baseline variable.** Using the SMD table,
   which variable shows the largest imbalance between groups in your simulated
   data? Is the SMD above or below 0.1?

4. **Why does the SRQ-20 show group imbalance?** Think back to
   [simulation-refinements.md](../simulation-refinements.md). The intervention
   group started with a higher average SRQ-20 score than the
   control group. Recall the discussion of how this imbalance affects the raw
   follow-up comparison. What does this suggest about the importance of the
   ANCOVA adjustment in Module 5?

---

## Summary

In this module you:

- Computed univariate summaries (mean, SD, n, %) for all baseline variables.
- Built a formatted Table 1 using `gtsummary`.
- Confirmed that randomization produced roughly balanced groups.
- Learned why significance tests of baseline differences are not appropriate in
  a randomized trial, and introduced standardized mean differences as a better
  measure of balance.

**Next:** Module 3 will examine the distributions of the primary and secondary
outcomes visually, using histograms, box plots, and individual-change plots.

---

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
