---
title: "Module 3: Data Visualization"
subtitle: "CRP 241 — IVAM-ED Teaching Case"
author: "Steve Grambow, PhD"
date: today
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Visualization is not decoration — it is analysis. Plots reveal distributional
features (skewness, outliers, floor effects), patterns across groups, and
individual variability that summary statistics conceal.

By the end of this module you should be able to:

1. Plot the distribution of a continuous variable with a histogram and density curve.
2. Compare distributions between two groups with side-by-side box plots.
3. Visualize change from baseline using an individual-trajectory plot.
4. Describe what each plot tells you about the data that a table cannot.

---------------------------------------------------------------------
---

## Setup

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(patchwork)  # combines multiple ggplot panels into one figure
```

```{r}
#| label: load-data

dat <- readRDS("../data/ivam_working.rds")

# Completers only (those with follow-up data) — used in several plots
completers <- filter(dat, missing_12wk == 0)
```

---------------------------------------------------------------------
---

## 1. Distribution of the Primary Outcome at Baseline

The SRQ-20 is the primary outcome: a 20-item mental-distress questionnaire.
Before looking at treatment effects, we should understand what its distribution
looks like in this sample.

```{r}
#| label: srq20-baseline-hist
#| fig-width: 7
#| fig-height: 4

ggplot(dat, aes(x = srq20_0)) +
  geom_histogram(binwidth = 1, fill = "#4472C4", color = "white", alpha = 0.85) +
  geom_density(aes(y = after_stat(count)), color = "#C00000", linewidth = 0.8,
               adjust = 1.2) +
  scale_x_continuous(breaks = seq(0, 20, 2), limits = c(-0.5, 20.5)) +
  labs(
    title    = "Distribution of SRQ-20 Score at Baseline",
    subtitle = "All 112 enrolled participants",
    x        = "SRQ-20 score (0 = no distress, 20 = maximum distress)",
    y        = "Number of participants",
    caption  = "Red curve: smoothed density overlay"
  ) +
  theme_minimal(base_size = 12)
```

::: {.callout-note}
**What to notice:**

- The distribution is right-skewed: most participants have low scores but a
  tail of participants have scores above 10.
- There is no sharp floor at 0 (the distribution is not dominated by zeros),
  meaning mental distress is present at a detectable level throughout this sample.
- The SRQ-20 is an integer sum score — that is why the bars fall at whole numbers.
:::

---------------------------------------------------------------------
---

## 2. Baseline Distributions by Group

Do the two groups look similar at baseline? We would hope so — randomization
should produce roughly equivalent groups, but it does not guarantee perfect
balance in every variable.

```{r}
#| label: baseline-by-group
#| fig-width: 8
#| fig-height: 9

# Reshape to long format so we can facet across all five outcomes at once
outcomes_long <- dat |>
  select(participant_id, group_label, srq20_0, sf36_0, sci_r_0, pss_0, hba1c_0) |>
  tidyr::pivot_longer(
    cols      = c(srq20_0, sf36_0, sci_r_0, pss_0, hba1c_0),
    names_to  = "outcome",
    values_to = "score"
  ) |>
  mutate(outcome = recode(outcome,
    srq20_0 = "SRQ-20\n(0–20, higher = worse distress)",
    sf36_0  = "SF-36\n(0–100, higher = better QoL)",
    sci_r_0 = "SCI-R\n(0–56, higher = better self-care)",
    pss_0   = "PSS\n(0–40, higher = more stress)",
    hba1c_0 = "HbA1c %\n(lower = better glycemic control)"
  ))

ggplot(outcomes_long, aes(x = group_label, y = score, fill = group_label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 1.5,
               width = 0.5) +
  geom_jitter(width = 0.12, alpha = 0.3, size = 0.9) +
  scale_fill_manual(values = c("Usual care" = "#ED7D31", "Smart speaker" = "#4472C4")) +
  facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
  labs(
    title   = "Baseline Outcome Distributions by Randomized Group",
    x       = NULL,
    y       = "Score",
    fill    = NULL,
    caption = "Points are individual participants; box shows median and IQR"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 9))
```

::: {.callout-note}
**What to notice:**

- The groups look broadly similar across all five outcomes — this is the expected
  result of randomization.
- The SRQ-20 box for the smart speaker group is shifted slightly upward relative
  to the usual care group (the intervention group started with slightly higher
  distress). This is chance imbalance; the ANCOVA adjustment in Module 5
  will account for it.
- The jittered individual points show the spread within each group. Box plots
  can hide multimodal distributions or extreme outliers that individual points
  reveal.
:::

---------------------------------------------------------------------
---

## 3. Change from Baseline: Individual Trajectories

The most informative visualization for a pre-post trial is an individual
trajectory plot: one line per participant, connecting their baseline score to
their follow-up score. This shows both the direction and magnitude of change
for each person, not just the group average.

```{r}
#| label: individual-change-srq20
#| fig-width: 8
#| fig-height: 5

# Reshape to long: one row per participant per time point
srq20_long <- completers |>
  select(participant_id, group_label, srq20_0, srq20_12) |>
  tidyr::pivot_longer(
    cols      = c(srq20_0, srq20_12),
    names_to  = "timepoint",
    values_to = "srq20"
  ) |>
  mutate(
    timepoint = recode(timepoint,
      srq20_0  = "Baseline\n(Week −1)",
      srq20_12 = "Follow-up\n(Week 12)"
    ),
    timepoint = factor(timepoint,
      levels = c("Baseline\n(Week −1)", "Follow-up\n(Week 12)"))
  )

ggplot(srq20_long,
       aes(x = timepoint, y = srq20, group = participant_id, color = group_label)) +
  geom_line(alpha = 0.3, linewidth = 0.5) +
  geom_point(alpha = 0.4, size = 1.2) +
  # Add group mean lines on top
  stat_summary(aes(group = group_label), fun = mean,
               geom = "line", linewidth = 1.5, alpha = 0.9) +
  stat_summary(aes(group = group_label), fun = mean,
               geom = "point", size = 3, shape = 18) +
  scale_color_manual(values = c("Usual care" = "#ED7D31", "Smart speaker" = "#4472C4")) +
  scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, 4)) +
  labs(
    title    = "Individual SRQ-20 Trajectories: Baseline to 12 Weeks",
    subtitle = "Thin lines = individual participants; thick lines = group means",
    x        = NULL,
    y        = "SRQ-20 score (higher = more distress)",
    color    = NULL,
    caption  = "Completers only (n = 103)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

::: {.callout-note}
**What to notice:**

- **Color key:** orange lines = Usual care group; blue lines = Smart speaker
  (intervention) group. The thick lines are group means; thin lines are individual
  participants.
- Both group means *decline* from baseline to follow-up — a drop in SRQ-20
  means *less* distress, which is an improvement.
- The smart speaker group mean (blue thick line) drops further than the usual care
  group mean (orange thick line). The vertical distance between the two thick lines
  at week 12 represents the unadjusted mean difference.
- Individual variability is large: some participants in both groups get worse,
  not better. The group-level treatment effect is a central tendency, not a
  guarantee for any individual.
- The crossing thin lines reveal that participants who started with high scores tend
  to show larger decreases — this is **regression to the mean**, which the ANCOVA
  model in Module 5 will address.
:::

---------------------------------------------------------------------
---

## 4. Comparing All Outcomes at Follow-Up

A side-by-side comparison of follow-up distributions across all outcomes gives
an overview of where the intervention had an effect and where it did not.

```{r}
#| label: followup-all-outcomes
#| fig-width: 8
#| fig-height: 9

followup_long <- completers |>
  select(participant_id, group_label,
         srq20_12, sf36_12, sci_r_12, pss_12, hba1c_12) |>
  tidyr::pivot_longer(
    cols      = c(srq20_12, sf36_12, sci_r_12, pss_12, hba1c_12),
    names_to  = "outcome",
    values_to = "score"
  ) |>
  mutate(outcome = recode(outcome,
    srq20_12 = "SRQ-20 (primary)\nhigher = worse distress",
    sf36_12  = "SF-36\nhigher = better QoL",
    sci_r_12 = "SCI-R\nhigher = better self-care",
    pss_12   = "PSS\nhigher = more stress",
    hba1c_12 = "HbA1c %\nlower = better glycemic control"
  ))

ggplot(followup_long, aes(x = group_label, y = score, fill = group_label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 1.5, width = 0.5) +
  geom_jitter(width = 0.12, alpha = 0.3, size = 0.9) +
  scale_fill_manual(values = c("Usual care" = "#ED7D31", "Smart speaker" = "#4472C4")) +
  facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
  labs(
    title   = "Outcome Distributions at 12-Week Follow-Up by Group",
    x       = NULL,
    y       = "Score",
    fill    = NULL,
    caption = "Points are individual participants; box shows median and IQR"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 9))
```

::: {.callout-note}
**Direction guide for comparing groups:**

- **SRQ-20:** Smart speaker group should be *lower* (less distress is better).
- **SF-36 and SCI-R:** Smart speaker group should be *higher* (more is better).
- **PSS:** Smart speaker group should be *lower* (less stress is better).
- **HbA1c:** Smart speaker group should be *lower* (lower % is better glycemic control).

The published trial found statistically significant effects on SRQ-20, SF-36,
SCI-R, and HbA1c, but not on PSS (the CI for PSS crossed zero). Look at your
plots: do any of the comparisons look visually similar (no clear separation),
even if a difference exists?
:::

---------------------------------------------------------------------
---

## 5. Floor and Ceiling Effects

A **floor effect** occurs when many participants score at or near the lowest
possible value — they cannot get lower, so the instrument cannot detect further
improvement. A **ceiling effect** is the reverse. Either distorts the analysis.

```{r}
#| label: floor-ceiling
#| fig-width: 8
#| fig-height: 4

p_srq <- ggplot(completers, aes(x = srq20_12, fill = group_label)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.65) +
  scale_x_continuous(limits = c(-0.5, 20.5), breaks = seq(0, 20, 4)) +
  scale_fill_manual(values = c("Usual care" = "#ED7D31", "Smart speaker" = "#4472C4")) +
  labs(title = "SRQ-20 at 12 weeks", x = "Score", y = "Count", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_sf36 <- ggplot(completers, aes(x = sf36_12, fill = group_label)) +
  geom_histogram(binwidth = 5, position = "identity", alpha = 0.65) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  scale_fill_manual(values = c("Usual care" = "#ED7D31", "Smart speaker" = "#4472C4")) +
  labs(title = "SF-36 at 12 weeks", x = "Score", y = "Count", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_srq + p_sf36 +
  plot_annotation(
    title   = "Follow-Up Distributions: Checking for Floor and Ceiling Effects",
    caption = "A pile-up of scores at 0 (SRQ-20) or 100 (SF-36) would indicate a floor or ceiling effect"
  )
```

::: {.callout-note}
**What to look for:** Are scores piled up against the boundary of the scale?
In the IVAM-ED data, mean SRQ-20 at follow-up is around 7, well above the floor
of 0, so floor effects are not a major concern. The SF-36 mean is around 55,
well below the ceiling of 100.
:::

---------------------------------------------------------------------
---

## 6. Practice Exercises

1. **Change direction.** For each of the five outcomes, write one sentence
   stating whether you expect the smart speaker group to have a *higher* or
   *lower* mean than the usual care group at follow-up, and why.

2. **Regression to the mean.** In the individual trajectory plot (Section 3),
   look at participants who started with SRQ-20 scores above 15. What tends to
   happen to their scores at follow-up? Does this happen in both groups? What
   does this tell you about interpreting within-group change?

3. **Plot the PSS trajectories.** Adapt the code from Section 3 to make an
   individual trajectory plot for the PSS (perceived stress) outcome. Does the
   visual separation between group means at follow-up look larger or smaller
   than for the SRQ-20? This anticipates the Module 4 finding that the PSS
   result was not statistically significant.

4. **HbA1c change.** Make a scatter plot of baseline HbA1c (`hba1c_0`) on the
   x-axis against 12-week HbA1c (`hba1c_12`) on the y-axis, with points colored
   by group. Add a reference line for y = x (no change). Which points are above
   the line (got worse) and which are below (improved)?

---------------------------------------------------------------------
---

## Summary

In this module you:

- Described the shape and spread of the SRQ-20 baseline distribution (right-skewed,
  no floor effect).
- Compared baseline and follow-up distributions between the two groups with box
  plots and individual trajectory plots.
- Identified the direction of expected treatment effects for each outcome.
- Introduced floor and ceiling effects as a data-quality check.
- Observed regression to the mean in the trajectory plot — a key motivation for
  the ANCOVA model in Module 5.

**Next:** Module 4 will quantify what the plots show: group means, mean changes
from baseline, 95% confidence intervals, and hypothesis tests.

---------------------------------------------------------------------
---

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
