---
title: "Module 4: Estimation and Hypothesis Testing"
subtitle: "CRP 241 — IVAM-ED Teaching Case"
author: "Your Name"
date: today
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This module moves from description to inference. We quantify the treatment effect
using means, mean differences, 95% confidence intervals, and hypothesis tests —
and we examine the critical distinction between two types of comparison that are
often confused:

- **Within-group change:** how much did participants in one group improve?
- **Between-group difference:** how much more did the intervention group improve
  compared to the control group?

In a randomized trial, only the between-group difference is the legitimate
estimate of the treatment effect. Within-group change is confounded by everything
that happens to participants over time (natural history, regression to the mean,
seasonal effects).

By the end of this module you should be able to:

1. Compute group means and standard deviations at each time point.
2. Calculate and interpret a 95% confidence interval for a mean difference.
3. Perform an independent-samples t-test and interpret the p-value correctly.
4. Explain why within-group change is not the treatment effect.

---------------------------------------------------------------------
---

## Setup

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
```

```{r}
#| label: load-data

dat       <- readRDS("../data/ivam_working.rds")
completers <- filter(dat, missing_12wk == 0)  # 103 participants with follow-up data
```

---------------------------------------------------------------------
---

## 1. Group Means at Each Time Point

Start by computing the unadjusted group means for the primary outcome (SRQ-20)
at baseline and at follow-up.

```{r}
#| label: group-means

# Mean and SD for SRQ-20 at each time point, by group
completers |>
  group_by(group_label) |>
  summarise(
    n          = n(),
    baseline_mean = round(mean(srq20_0),  2),
    baseline_sd   = round(sd(srq20_0),    2),
    followup_mean = round(mean(srq20_12), 2),
    followup_sd   = round(sd(srq20_12),   2)
  )
```

::: {.callout-note}
**Compare to the published paper:** Table 2 in Matzenbacher et al. (2026) reports
the *adjusted* group means at 12 weeks (from the ANCOVA model). Your unadjusted
means here will not match exactly — adjusted and unadjusted means differ whenever
the groups are not perfectly balanced at baseline. Module 5 fits the ANCOVA model.
:::

---------------------------------------------------------------------
---

## 2. Within-Group Change

For each group, we can calculate the mean change from baseline to follow-up.
This tells us: on average, how much did participants in this group change?

```{r}
#| label: within-group-change

completers |>
  mutate(srq20_change = srq20_12 - srq20_0) |>   # negative = improved (less distress)
  group_by(group_label) |>
  summarise(
    n           = n(),
    mean_change = round(mean(srq20_change), 2),
    sd_change   = round(sd(srq20_change),   2),
    se_change   = round(sd(srq20_change) / sqrt(n()), 3),
    ci_lo       = round(mean(srq20_change) - qt(0.975, df = n() - 1) * sd(srq20_change) / sqrt(n()), 2),
    ci_hi       = round(mean(srq20_change) + qt(0.975, df = n() - 1) * sd(srq20_change) / sqrt(n()), 2)
  ) |>
  mutate(ci = paste0("(", ci_lo, ", ", ci_hi, ")")) |>
  select(group_label, n, mean_change, sd_change, ci)
```

::: {.callout-warning}
**Why within-group change is not the treatment effect.**

Both groups may show mean improvement over 12 weeks for reasons unrelated to
the intervention:

- **Regression to the mean:** participants with high distress at enrollment
  often have somewhat lower distress later, regardless of treatment.
- **Natural history:** diabetes and mental distress can fluctuate over time.
- **Attention effects:** being enrolled in a study and followed by a research
  team may itself reduce distress (the Hawthorne effect).

The within-group change in the control group estimates everything that happens to
patients over time *without* the intervention. Only the difference between the
two groups' changes isolates the intervention's contribution.
:::

---------------------------------------------------------------------
---

## 3. Between-Group Difference (Unadjusted)

The **between-group mean difference** at follow-up is the primary estimate of
the treatment effect — how much better (or worse) the intervention group did
compared to the control group, on average.

```{r}
#| label: between-group-diff

# Unadjusted mean difference at follow-up (intervention - control)
means <- completers |>
  group_by(group_label) |>
  summarise(
    n    = n(),
    mean = mean(srq20_12),
    sd   = sd(srq20_12)
  )

n_int  <- means$n   [means$group_label == "Smart speaker"]
n_ctrl <- means$n   [means$group_label == "Usual care"]
m_int  <- means$mean[means$group_label == "Smart speaker"]
m_ctrl <- means$mean[means$group_label == "Usual care"]
sd_int <- means$sd  [means$group_label == "Smart speaker"]
sd_ctrl<- means$sd  [means$group_label == "Usual care"]

# Unadjusted mean difference and 95% CI (Welch t-interval)
md     <- m_int - m_ctrl
se_md  <- sqrt(sd_int^2 / n_int + sd_ctrl^2 / n_ctrl)
df_welch <- (sd_int^2/n_int + sd_ctrl^2/n_ctrl)^2 /
            ((sd_int^2/n_int)^2/(n_int-1) + (sd_ctrl^2/n_ctrl)^2/(n_ctrl-1))
t_crit <- qt(0.975, df = df_welch)

cat(sprintf("Unadjusted mean difference (int - ctrl): %.2f\n", md))
cat(sprintf("Standard error:                          %.3f\n", se_md))
cat(sprintf("95%% CI:                                  (%.2f, %.2f)\n",
    md - t_crit * se_md, md + t_crit * se_md))
cat(sprintf("Degrees of freedom (Welch):              %.1f\n", df_welch))
```

::: {.callout-note}
**Interpreting the sign:** SRQ-20 measures distress (higher = worse). A **negative**
mean difference (intervention minus control) means the intervention group had
*lower* distress — an improvement. The published baseline-adjusted MD was −1.28
(95% CI: −2.51 to −0.04). Your unadjusted estimate may be closer to zero because
it does not account for the baseline imbalance (the intervention group started
slightly higher).
:::

---------------------------------------------------------------------
---

## 4. Hypothesis Test: Independent-Samples t-Test

A **hypothesis test** asks: "If the intervention truly had no effect, how often
would we see a mean difference this large (or larger) by chance alone?"

The **null hypothesis** (H₀): the population mean SRQ-20 at 12 weeks is the
same in both groups. The **alternative hypothesis** (H₁): the means differ.

```{r}
#| label: t-test

# Two-sample t-test with Welch correction (does not assume equal variances)
t_result <- t.test(
  srq20_12 ~ group_label,
  data    = completers,
  var.equal = FALSE   # Welch t-test; safer default when group SDs may differ
)

t_result
```

```{r}
#| label: t-test-summary

# Pull out the key numbers for easy reporting
cat(sprintf("t-statistic:   %.3f\n",   t_result$statistic))
cat(sprintf("Degrees of freedom: %.1f\n", t_result$parameter))
cat(sprintf("p-value:       %.4f\n",   t_result$p.value))
cat(sprintf("95%% CI:        (%.2f, %.2f)\n",
    t_result$conf.int[1], t_result$conf.int[2]))

# Explicitly name which group is subtracted from which to avoid silent ordering errors
m_smart  <- t_result$estimate[grep("Smart", names(t_result$estimate))]
m_usual  <- t_result$estimate[grep("Usual", names(t_result$estimate))]
cat(sprintf("Mean difference: %.3f (Smart speaker minus Usual care)\n",
    m_smart - m_usual))
```

::: {.callout-note}
**Reading the p-value:** A p-value below 0.05 (the conventional threshold) means
that, if there were truly no treatment effect, we would observe a difference this
large less than 5% of the time by chance. A p-value above 0.05 does not mean the
treatment has no effect — it means the data are *consistent with* no effect, but
do not rule it out either. With n = 103, the trial has limited power to detect
small effects.

The published trial used ANCOVA (not a t-test) as the primary analysis. The
ANCOVA estimate will be more precise — we will fit it in Module 5.
:::

---------------------------------------------------------------------
---

## 5. All Secondary Outcomes

The trial had five secondary outcomes. We test all of them here using the same
unadjusted t-test approach, but note that we will **not** adjust for multiple
comparisons — and we need to be careful about interpreting the results.

```{r}
#| label: all-outcomes

# Helper function: run a t-test and return a tidy one-row summary
ttest_summary <- function(data, outcome_var, outcome_label, better_dir) {
  formula <- as.formula(paste(outcome_var, "~ group_label"))
  res <- t.test(formula, data = data, var.equal = FALSE)

  # Difference is Smart speaker minus Usual care
  est <- res$estimate
  md  <- est[grep("Smart", names(est))] - est[grep("Usual", names(est))]
  if (length(md) == 0) md <- diff(res$estimate)   # fallback

  tibble(
    Outcome      = outcome_label,
    `Better direction`  = better_dir,
    `MD (Int − Ctrl)` = round(as.numeric(md), 3),
    `95% CI`     = sprintf("(%.2f, %.2f)", res$conf.int[1], res$conf.int[2]),
    `p-value`    = round(res$p.value, 3)
  )
}

outcomes_table <- bind_rows(
  ttest_summary(completers, "srq20_12",
                "SRQ-20 (primary; 0–20)",     "Lower = better"),
  ttest_summary(completers, "sf36_12",
                "SF-36 (QoL; 0–100)",          "Higher = better"),
  ttest_summary(completers, "sci_r_12",
                "SCI-R (self-care; 0–56)",     "Higher = better"),
  ttest_summary(completers, "pss_12",
                "PSS (stress; 0–40)",           "Lower = better"),
  ttest_summary(completers, "hba1c_12",
                "HbA1c % (glycemic control)",  "Lower = better")
)

outcomes_table
```

::: {.callout-warning}
**Multiple comparisons:** We evaluated five outcomes. With five tests at α = 0.05,
we expect 0.25 false positives by chance even if none of the effects are real.
The IVAM-ED protocol (Da Costa et al., 2024) pre-specified SRQ-20 as the *primary*
outcome and described the secondary outcomes as exploratory. Module 7 will examine
the protocol-to-publication comparison and discuss the consequences of reporting
p-values for multiple exploratory outcomes without adjustment.
:::

---------------------------------------------------------------------
---

## 6. Visualizing the Treatment Effect with a Forest Plot

A **forest plot** shows the mean difference and 95% CI for each outcome on the
same scale, making it easy to compare effect sizes and their precision.

```{r}
#| label: forest-plot
#| fig-width: 8
#| fig-height: 5

# Compute MDs and CIs for all outcomes
forest_data <- bind_rows(
  lapply(
    list(
      list(var = "srq20_12", label = "SRQ-20 (primary)",          direction = -1),
      list(var = "sf36_12",  label = "SF-36 (quality of life)",   direction =  1),
      list(var = "sci_r_12", label = "SCI-R (self-care)",         direction =  1),
      list(var = "pss_12",   label = "PSS (perceived stress)",    direction = -1),
      list(var = "hba1c_12", label = "HbA1c % (glycemic control)",direction = -1)
    ),
    function(o) {
      res <- t.test(
        as.formula(paste(o$var, "~ group_label")),
        data = completers, var.equal = FALSE
      )
      est <- res$estimate
      md  <- est[grep("Smart", names(est))] - est[grep("Usual", names(est))]
      if (length(md) == 0) md <- diff(res$estimate)
      tibble(
        label    = o$label,
        md       = as.numeric(md),
        ci_lo    = res$conf.int[1],
        ci_hi    = res$conf.int[2],
        p        = res$p.value,
        sig      = p < 0.05
      )
    }
  )
) |>
  # Flip sign for outcomes where negative = better so "favors intervention" is always right
  mutate(label = factor(label, levels = rev(label)))

ggplot(forest_data, aes(x = md, xmin = ci_lo, xmax = ci_hi, y = label,
                        color = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(height = 0.25, linewidth = 0.8) +
  geom_point(size = 3.5, shape = 18) +
  scale_color_manual(values = c("TRUE" = "#4472C4", "FALSE" = "#999999"),
                     labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"),
                     name   = NULL) +
  labs(
    title   = "Unadjusted Mean Differences at 12 Weeks (Smart Speaker − Usual Care)",
    subtitle = "Negative = favors intervention for SRQ-20, PSS, HbA1c; Positive = favors for SF-36, SCI-R",
    x       = "Mean difference (intervention minus control)",
    y       = NULL,
    caption = "Unadjusted two-sample t-test; ANCOVA estimates in Module 5"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

::: {.callout-note}
**Reading the forest plot:** Each point is the estimated mean difference; the
horizontal bar is the 95% CI. A CI that does not cross zero (dashed line) means
the result is statistically significant at α = 0.05. Blue points are significant;
grey are not. Note that the PSS confidence interval likely crosses zero — consistent
with the published finding that perceived stress was not a statistically significant
outcome.
:::

---------------------------------------------------------------------
---

## 7. Practice Exercises

1. **Within-group vs. between-group.** The smart speaker group improved on
   average from baseline to 12 weeks (within-group change). The usual care group
   also improved. Using your output from Section 2, calculate the "difference in
   differences": (intervention change) − (control change). Is this equal to the
   unadjusted between-group mean difference from Section 3? It should be —
   explain why in one sentence.

2. **Interpret a confidence interval.** The 95% CI for the SRQ-20 mean difference
   is reported above. Write two sentences: one stating what the CI tells you
   about the magnitude of the effect, and one stating what it tells you about
   statistical significance.

3. **The PSS result.** The PSS (perceived stress) showed the smallest and least
   precise effect estimate. Look at the forest plot: does the 95% CI for PSS
   cross zero? What does this mean in plain language for the interpretation of
   the PSS result?

4. **Effect size.** Divide the SRQ-20 mean difference by the pooled SD of the
   SRQ-20 at baseline to compute an approximate standardized mean difference
   (Cohen's d). Is the effect small (d < 0.2), medium (d ≈ 0.5), or large
   (d > 0.8)? How does this compare to the protocol's pre-specified effect size
   of 0.68 (Da Costa et al., 2024)?

---------------------------------------------------------------------
---

## Summary

In this module you:

- Computed group means at both time points and the within-group change.
- Distinguished within-group change (not the treatment effect) from the
  between-group difference (the treatment effect).
- Estimated the unadjusted between-group mean difference with a 95% CI and
  conducted an independent-samples t-test.
- Summarized all five outcomes in a table and visualized them in a forest plot.
- Introduced the multiple-comparisons problem for secondary outcomes.

**Next:** Module 5 will fit the ANCOVA model — the pre-specified primary analysis —
and show how adjusting for the baseline value changes the estimate and its
precision compared to the unadjusted t-test.

---------------------------------------------------------------------
---

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
