---
title: "Module 5: Linear Regression and ANCOVA"
subtitle: "CRP 241 — IVAM-ED Teaching Case"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: flatly
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

In Module 4 we used an unadjusted t-test to compare groups at follow-up. The
pre-specified primary analysis in the IVAM-ED protocol used a different approach:
**ANCOVA (Analysis of Covariance)** — a linear regression model that includes
the baseline value of the outcome as a covariate alongside the treatment group
indicator.

By the end of this module you should be able to:

1. Explain why adjusting for baseline improves precision without introducing bias.
2. Fit a baseline-adjusted ANCOVA model using `lm()` and interpret the treatment
   coefficient.
3. Fit the fully-adjusted model with additional covariates and compare it to the
   baseline-adjusted model.
4. Report results in the format used in the published paper (Table 2).
5. Understand what "adjusted mean" means and how it differs from an observed mean.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
```

```{r}
#| label: load-data

dat        <- readRDS("../data/ivam_working.rds")
completers <- filter(dat, missing_12wk == 0)   # 103 participants
```

------------------------------------------------------------------------

## 1. Why Adjust for Baseline?

When participants are randomized, the two groups are *expected* to be balanced at
baseline on average — but in any finite sample, chance imbalances occur. In our
simulated data (and the real IVAM-ED trial), the intervention group started with
a slightly higher SRQ-20 score than the control group.

**Two consequences of baseline imbalance:**

1. **The unadjusted mean difference is biased.** If the intervention group starts
   higher, regression to the mean will push them down more, even without a real
   treatment effect. Adjusting for baseline removes this confounding.

2. **Adjusting for baseline increases precision.** Because baseline scores are
   strongly correlated with follow-up scores, including the baseline as a covariate
   reduces residual variance. This makes confidence intervals narrower and tests
   more powerful — for free, without collecting additional data.

```{r}
#| label: baseline-correlation

# How strongly correlated are baseline and follow-up SRQ-20?
with(completers, round(cor(srq20_0, srq20_12, use = "complete.obs"), 3))
```

::: {.callout-note}
**Interpreting the correlation:** A value near 0.7 means that knowing a
participant's baseline score explains about 49% of the variance in their
follow-up score (r² = 0.70² ≈ 0.49). Including the baseline in the regression
model can therefore reduce the residual variance by roughly half, which substantially
narrows the confidence interval for the treatment effect.
:::

------------------------------------------------------------------------

## 2. The Baseline-Adjusted ANCOVA Model

The baseline-adjusted ANCOVA model includes two predictors:
- `group`: the randomized treatment assignment (1 = intervention, 0 = control)
- `srq20_0`: the participant's SRQ-20 score at baseline

```{r}
#| label: ancova-baseline

# Baseline-adjusted model: y_12 ~ group + y_0
model_base <- lm(srq20_12 ~ group + srq20_0, data = completers)
summary(model_base)
```

```{r}
#| label: ancova-base-results

# Extract the key estimates in a clean format
coef_base <- summary(model_base)$coefficients
ci_base   <- confint(model_base)

results_base <- data.frame(
  Term         = rownames(coef_base),
  Estimate     = round(coef_base[, "Estimate"], 3),
  SE           = round(coef_base[, "Std. Error"], 3),
  `95% CI`     = paste0("(", round(ci_base[, 1], 2), ", ", round(ci_base[, 2], 2), ")"),
  `p-value`    = round(coef_base[, "Pr(>|t|)"], 4),
  check.names  = FALSE
)

results_base
```

::: {.callout-note}
**Interpreting the `group` coefficient (the treatment effect):**

- This is the estimated mean difference in SRQ-20 at 12 weeks between the
  intervention and control groups, **holding baseline SRQ-20 constant**.
- A negative value means the intervention group had lower distress (better outcome).
- Compare this to the unadjusted mean difference from Module 4. The ANCOVA
  estimate should be larger in magnitude because it corrects for the baseline
  imbalance (intervention group started higher).
- The 95% CI should be narrower than the unadjusted CI, reflecting the increased
  precision from including the baseline covariate.

**Published baseline-adjusted MD:** −1.28 (95% CI: −2.51 to −0.04). Your
estimate will be close but not identical because this is a simulation.
:::

------------------------------------------------------------------------

## 3. Adjusted Means

The ANCOVA model produces **adjusted means** — the estimated group means at
follow-up after accounting for baseline differences. These are what the paper
reports in Table 2 as "Adjusted mean (SE)".

The adjusted means are evaluated at the **grand mean** of the baseline covariate
(not the group-specific mean). This removes the effect of any baseline imbalance.

```{r}
#| label: adjusted-means

# Grand mean of baseline SRQ-20 across all completers
grand_mean_srq20 <- mean(completers$srq20_0)
cat("Grand mean of baseline SRQ-20:", round(grand_mean_srq20, 2), "\n\n")

# Adjusted means: predict at grand mean of baseline, for each group
adj_means <- predict(
  model_base,
  newdata = data.frame(
    group   = c(0, 1),
    srq20_0 = c(grand_mean_srq20, grand_mean_srq20)
  ),
  se.fit = TRUE
)

cat("Adjusted mean (SE) — Usual care:    ",
    round(adj_means$fit[1], 2), "(", round(adj_means$se.fit[1], 2), ")\n")
cat("Adjusted mean (SE) — Smart speaker: ",
    round(adj_means$fit[2], 2), "(", round(adj_means$se.fit[2], 2), ")\n")
cat("\nPublished (Table 2): Usual care 7.66 (0.43), Smart speaker 6.39 (0.45)\n")
```

::: {.callout-note}
**Why adjusted means differ from raw means:** The raw follow-up means reflect
which participants ended up in each group, including any baseline imbalance.
The adjusted means represent what the expected follow-up score would be for a
"typical" participant (at the grand mean of the baseline) in each group.
:::

------------------------------------------------------------------------

## 4. The Fully Adjusted Model

The protocol specified a **fully adjusted** model that additionally controls for:
age, sex, years of education, individual income, and MMSE score. These variables
are included to account for any remaining between-group differences that could
influence the outcome.

```{r}
#| label: ancova-full

# Fully adjusted model: adds age, sex, education, income, MMSE as covariates
model_full <- lm(
  srq20_12 ~ group + srq20_0 + age + female + edu_years + income_wages + mmse_0,
  data = completers
)
summary(model_full)
```

```{r}
#| label: full-results

coef_full <- summary(model_full)$coefficients
ci_full   <- confint(model_full)

data.frame(
  Term      = rownames(coef_full),
  Estimate  = round(coef_full[, "Estimate"], 3),
  SE        = round(coef_full[, "Std. Error"], 3),
  `95% CI`  = paste0("(", round(ci_full[, 1], 2), ", ", round(ci_full[, 2], 2), ")"),
  `p-value` = round(coef_full[, "Pr(>|t|)"], 4),
  check.names = FALSE
)
```

::: {.callout-note}
**Comparing baseline-adjusted vs. fully adjusted:**

- The treatment effect (`group` coefficient) may change slightly between the two
  models. In the published paper, the fully adjusted MD was −1.46 (95% CI: −2.73
  to −0.19) versus −1.28 baseline-adjusted — slightly larger after full adjustment.
- Adding more covariates can either increase or decrease the treatment estimate
  depending on whether those covariates are positively or negatively correlated
  with both group assignment and the outcome.
- The primary advantage of the fully adjusted model is further precision gain
  when the additional covariates explain residual variance.
:::

------------------------------------------------------------------------

## 5. Replicating Table 2: All Outcomes

The paper's Table 2 reports baseline-adjusted and fully adjusted results for all
five outcomes. Here we fit the baseline-adjusted model for each and compile the
results.

```{r}
#| label: table2-replication

# Helper: fit ANCOVA for one outcome and return a tidy summary row
ancova_row <- function(data, outcome_12, outcome_0, label) {
  fmla <- as.formula(paste(outcome_12, "~", "group +", outcome_0))
  m    <- lm(fmla, data = data)
  ci   <- confint(m)["group", ]
  coef_row <- coef(summary(m))["group", ]

  data.frame(
    Outcome       = label,
    MD            = round(coef_row["Estimate"], 2),
    SE            = round(coef_row["Std. Error"], 2),
    `CI lower`    = round(ci[1], 2),
    `CI upper`    = round(ci[2], 2),
    `p-value`     = round(coef_row["Pr(>|t|)"], 3),
    check.names   = FALSE
  )
}

table2 <- bind_rows(
  ancova_row(completers, "srq20_12", "srq20_0", "SRQ-20 (primary)"),
  ancova_row(completers, "sf36_12",  "sf36_0",  "SF-36 (quality of life)"),
  ancova_row(completers, "sci_r_12", "sci_r_0", "SCI-R (self-care)"),
  ancova_row(completers, "pss_12",   "pss_0",   "PSS (perceived stress)"),
  ancova_row(completers, "hba1c_12", "hba1c_0", "HbA1c %")
)

table2
```

```{r}
#| label: table2-published

# Published baseline-adjusted results (Table 2, Matzenbacher et al. 2026)
# for comparison
published <- data.frame(
  Outcome    = c("SRQ-20 (primary)", "SF-36 (quality of life)",
                 "SCI-R (self-care)", "PSS (perceived stress)", "HbA1c %"),
  MD_pub     = c(-1.28, 9.46, 3.40, -3.00, -0.48),
  CI_pub     = c("(-2.51, -0.04)", "(3.65, 15.26)", "(1.61, 5.19)",
                 "(-6.20, 0.20)", "(-0.85, -0.11)"),
  p_pub      = c(0.04, 0.001, "<0.001", 0.07, 0.01)
)

published
```

::: {.callout-note}
**Comparing your estimates to the published values:** Your estimates should have
the same sign and approximate magnitude as the published results. Exact matches
are not expected — the simulation approximates, but does not reconstruct, the
real data. The key patterns to confirm:

- SRQ-20: negative MD, p < 0.05 (significant)
- SF-36 and SCI-R: positive MD, p < 0.05 (significant)
- PSS: negative MD, p > 0.05 (not significant; CI should cross zero)
- HbA1c: negative MD, p < 0.05 (significant)
:::

------------------------------------------------------------------------

## 6. Visualizing Adjusted vs. Unadjusted Estimates

```{r}
#| label: compare-estimates
#| fig-width: 8
#| fig-height: 4

# Collect unadjusted and adjusted MDs for SRQ-20
unadj_t  <- t.test(srq20_12 ~ group_label, data = completers)
unadj_est <- diff(rev(unadj_t$estimate))   # Smart speaker - Usual care

unadj_ci <- unadj_t$conf.int
adj_ci   <- confint(model_base)["group", ]
adj_est  <- coef(model_base)["group"]

compare <- data.frame(
  Model    = factor(c("Unadjusted (t-test)", "Baseline-adjusted ANCOVA"),
                    levels = c("Unadjusted (t-test)", "Baseline-adjusted ANCOVA")),
  MD       = c(unadj_est, adj_est),
  lo       = c(unadj_ci[1], adj_ci[1]),
  hi       = c(unadj_ci[2], adj_ci[2])
)

ggplot(compare, aes(x = MD, xmin = lo, xmax = hi, y = Model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(height = 0.2, linewidth = 0.9) +
  geom_point(size = 4, shape = 18, color = "#4472C4") +
  annotate("point", x = -1.28, y = 2, size = 4, shape = 4,
           color = "#C00000") +
  annotate("text",  x = -1.28, y = 2.25, label = "Published (−1.28)",
           color = "#C00000", size = 3.2) +
  labs(
    title   = "SRQ-20: Unadjusted vs. Baseline-Adjusted Mean Difference",
    subtitle = "Red × = published baseline-adjusted estimate",
    x       = "Mean difference (Smart speaker − Usual care)",
    y       = NULL
  ) +
  theme_minimal(base_size = 12)
```

::: {.callout-note}
**What to notice:**

- The ANCOVA estimate should be larger in absolute value than the unadjusted
  estimate (the adjustment corrects for the baseline imbalance that attenuated
  the raw difference).
- The ANCOVA CI should be narrower (more precise), demonstrating the efficiency
  gain from baseline adjustment.
:::

------------------------------------------------------------------------

## 7. Practice Exercises

1. **Coefficient interpretation.** Write one sentence interpreting the `group`
   coefficient from the baseline-adjusted ANCOVA as you would in a results
   section of a paper.

2. **Baseline coefficient.** The `srq20_0` coefficient in the ANCOVA model is
   approximately 0.70. What does this tell you about the relationship between
   baseline and follow-up SRQ-20 scores?

3. **Adjusted means by hand.** Using the estimated coefficients from
   `model_base`, predict the adjusted mean SRQ-20 at 12 weeks for a participant
   in the control group (`group = 0`) with a baseline SRQ-20 equal to the grand
   mean. Does this match the output from Section 3?

4. **Degrees of freedom.** The baseline-adjusted model has 103 observations and
   estimates 3 coefficients (intercept, group, baseline). How many residual
   degrees of freedom does it have? Why does the fully adjusted model (7
   coefficients) have fewer? What is the practical consequence of having fewer
   degrees of freedom?

------------------------------------------------------------------------

## Summary

In this module you:

- Explained why ANCOVA is preferred over an unadjusted t-test in a randomized
  trial with continuous outcomes.
- Fitted the baseline-adjusted ANCOVA and interpreted the treatment coefficient
  and adjusted means.
- Extended to the fully adjusted model with additional covariates.
- Reproduced the structure of the published Table 2 for all five outcomes.
- Visualized the precision gain from baseline adjustment.

**Next:** Module 6 examines missing data — how the 9 missing participants affect
the analysis and what sensitivity analyses can tell us about robustness.

------------------------------------------------------------------------

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
