---
title: "Module 7: Interpretation and Critical Appraisal"
subtitle: "CRP 241 — IVAM-ED Teaching Case"
author: "Steve Grambow, PhD"
date: today
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Statistical analysis produces numbers — but numbers require interpretation.
This final module develops the critical-appraisal skills needed to read a
clinical trial report with appropriate skepticism.

We focus on four questions:

1. Is a statistically significant result also **clinically meaningful**?
2. What is the difference between a **primary** and an **exploratory** finding?
3. What happens to the interpretation when you test **many outcomes without
   multiplicity adjustment**?
4. When the **published results differ from the pre-specified protocol**,
   what questions should you ask?

By the end of this module you should be able to:

1. Distinguish statistical significance from clinical significance.
2. Identify which findings in a published paper are confirmatory versus exploratory.
3. Apply a simple Bonferroni correction and explain what changes.
4. Compare the published IVAM-ED analysis to its pre-registered protocol and
   identify at least three deviations.

---------------------------------------------------------------------
---

## Setup

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
```

```{r}
#| label: load-data

dat        <- readRDS("../data/ivam_working.rds")
completers <- filter(dat, missing_12wk == 0)
```

---------------------------------------------------------------------
---

## 1. Statistical Significance vs. Clinical Significance

A statistically significant result (p < 0.05) tells us the observed effect is
unlikely to be due to chance. It says nothing about whether the effect is large
enough to matter in practice.

Clinical significance asks: **Is this effect large enough to benefit patients in
a meaningful way?**

### 1a. The minimum clinically important difference (MCID)

For the SRQ-20, the MCID — the smallest change a patient would notice and
consider beneficial — is typically reported as **3 points** in comparable
populations. (Note: a formal MCID for this specific population may not be
published; this is an instructive approximation.)

```{r}
#| label: effect-size

# Fit baseline-adjusted ANCOVA for primary outcome
model <- lm(srq20_12 ~ group + srq20_0, data = completers)
md    <- coef(model)["group"]
ci    <- confint(model)["group", ]

# Pooled SD for standardized effect size
pooled_sd <- completers |>
  summarise(sd = sd(srq20_0)) |>
  pull(sd)

cohens_d  <- md / pooled_sd

cat("SRQ-20 primary outcome:\n")
cat(sprintf("  Baseline-adjusted MD:  %.2f points\n", md))
cat(sprintf("  95%% CI:                (%.2f, %.2f)\n", ci[1], ci[2]))
cat(sprintf("  Pooled baseline SD:    %.2f\n", pooled_sd))
cat(sprintf("  Cohen's d:             %.2f\n", cohens_d))
cat(sprintf("\n  MCID threshold (approximate): 3 points\n"))
cat(sprintf("  Does the MD exceed the MCID?  %s\n",
    ifelse(abs(md) >= 3, "Yes", "No — upper CI does not reach MCID")))
```

::: {.callout-warning}
**A statistically significant but clinically modest effect.** The adjusted MD
is approximately −1.2 to −1.5 points on a 20-point scale. The 95% CI's upper
bound is close to zero, and neither the point estimate nor the CI upper bound
reaches the approximate 3-point MCID. The result is statistically significant
but the clinical magnitude is modest.

This is a common pattern in trials with modest effect sizes and limited power:
the study is large enough to detect the effect as real but the effect itself
may not be large enough to change clinical practice on its own. The authors
acknowledged this limitation and noted the trial was designed with an effect
size of 0.68 — the observed effect may have been smaller.
:::

### 1b. Visualizing the CI against the MCID

```{r}
#| label: mcid-plot
#| fig-width: 7
#| fig-height: 2.8

mcid <- -3   # negative because lower SRQ-20 = better

ggplot(data.frame(MD = md, lo = ci[1], hi = ci[2]),
       aes(x = MD, xmin = lo, xmax = hi, y = 1)) +
  geom_rect(aes(xmin = -Inf, xmax = mcid, ymin = 0.5, ymax = 1.5),
            fill = "#E2EFDA", alpha = 0.5) +
  annotate("text", x = mcid - 0.1, y = 1.35,
           label = "MCID\nthreshold", hjust = 1, size = 3.2, color = "#538135") +
  geom_vline(xintercept = 0,    linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = mcid, linetype = "dotted", color = "#538135", linewidth = 0.8) +
  geom_errorbarh(height = 0.25, linewidth = 1) +
  geom_point(size = 4, shape = 18, color = "#4472C4") +
  scale_x_continuous(limits = c(-4, 1), breaks = seq(-4, 1, 1)) +
  scale_y_continuous(limits = c(0.3, 1.6)) +
  labs(
    title   = "SRQ-20 Treatment Effect vs. Approximate Minimum Clinically Important Difference",
    x       = "Mean difference (Smart speaker − Usual care)",
    y       = NULL,
    caption = "Green shaded region = effects that meet or exceed the approximate MCID of 3 points"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

---------------------------------------------------------------------
---

## 2. Primary vs. Exploratory Findings

The IVAM-ED protocol pre-specified **one primary outcome** (SRQ-20 at 12 weeks)
and **five secondary outcomes** (SF-36, SCI-R, PSS, HbA1c, and blood pressure).
Only the primary outcome was powered for confirmatory inference.

The secondary outcomes were described as **exploratory** because:
- The trial was not powered to detect effects on secondary outcomes.
- Multiple outcomes were tested without adjustment for multiple comparisons.
- Exploratory findings generate hypotheses for future trials; they do not confirm effects.

```{r}
#| label: primary-vs-exploratory

# Reproduce Table 2 from Module 5 and label the outcomes
ancova_row <- function(data, y12, y0, label, type) {
  m  <- lm(as.formula(paste(y12, "~ group +", y0)), data = data)
  ci <- confint(m)["group", ]
  coef_row <- coef(summary(m))["group", ]
  data.frame(
    Outcome = label, Type = type,
    MD      = round(coef_row["Estimate"], 2),
    `95% CI`= paste0("(", round(ci[1], 2), ", ", round(ci[2], 2), ")"),
    p       = round(coef_row["Pr(>|t|)"], 3),
    check.names = FALSE
  )
}

bind_rows(
  ancova_row(completers, "srq20_12", "srq20_0", "SRQ-20",  "Primary (confirmatory)"),
  ancova_row(completers, "sf36_12",  "sf36_0",  "SF-36",   "Secondary (exploratory)"),
  ancova_row(completers, "sci_r_12", "sci_r_0", "SCI-R",   "Secondary (exploratory)"),
  ancova_row(completers, "pss_12",   "pss_0",   "PSS",     "Secondary (exploratory)"),
  ancova_row(completers, "hba1c_12", "hba1c_0", "HbA1c %", "Secondary (exploratory)")
)
```

---------------------------------------------------------------------
---

## 3. Multiple Comparisons

When you test five outcomes at α = 0.05, the probability of getting at least
one false positive (by chance alone) is:

$$1 - (1 - 0.05)^5 = 1 - 0.95^5 \approx 0.226$$

That is, even if none of the interventions have a true effect on the secondary
outcomes, we expect about a 23% chance of seeing at least one p < 0.05 among
the five tests just by random variation.

### 3a. Bonferroni correction

The simplest multiplicity correction is **Bonferroni**: divide the target
α by the number of tests. With 5 tests, the corrected threshold is
0.05 / 5 = 0.01.

```{r}
#| label: bonferroni

results_all <- bind_rows(
  ancova_row(completers, "srq20_12", "srq20_0", "SRQ-20",  "Primary"),
  ancova_row(completers, "sf36_12",  "sf36_0",  "SF-36",   "Secondary"),
  ancova_row(completers, "sci_r_12", "sci_r_0", "SCI-R",   "Secondary"),
  ancova_row(completers, "pss_12",   "pss_0",   "PSS",     "Secondary"),
  ancova_row(completers, "hba1c_12", "hba1c_0", "HbA1c %", "Secondary")
)

results_all |>
  mutate(
    `Sig at α = 0.05`   = ifelse(p < 0.05,    "Yes", "No"),
    `Sig at α = 0.01\n(Bonferroni)` = ifelse(p < 0.01, "Yes", "No")
  )
```

::: {.callout-note}
**What the Bonferroni correction shows:** The primary outcome (SRQ-20) should
remain significant under the corrected threshold if p < 0.01. Some secondary
outcomes that were significant at 0.05 may no longer meet the stricter threshold.
This illustrates why exploratory secondary findings require cautious interpretation
and should be labeled as hypothesis-generating rather than confirmatory.
:::

---------------------------------------------------------------------
---

## 4. Protocol-to-Publication Comparison

One of the most important skills in evidence-based medicine is **comparing what
was done** (published paper) **to what was planned** (protocol and SAP).

The IVAM-ED protocol was published in *Trials* in 2024 (Da Costa et al.) before
the results were available. The results paper was published in *JAMA Network Open*
in 2026. Both are in `source/extracted/`.

### 4a. Changes found between the protocol and the publication

Below are four specific changes that a careful reader can identify by comparing
the two documents. Read the relevant sections in `source/extracted/` to verify each one.

```{r}
#| label: protocol-changes

# This code block is a structured record — not analysis output.
# It serves as a reference table for the discussion below.
changes <- data.frame(
  `Change #` = 1:4,
  Feature    = c(
    "Age subgroup cutoff",
    "Education subgroup category",
    "History of anxiety subgroup",
    "Secondary outcome p-values"
  ),
  `Protocol/SAP specified` = c(
    "Age ≥ 80 years as the cutoff for the older-age subgroup",
    "Graduate education (vs. lower) as the subgroup category",
    "Not pre-specified",
    "P-values for secondary outcomes would NOT be reported (exploratory)"
  ),
  `Published` = c(
    "Age ≥ 75 years — changed because too few participants were aged ≥ 80",
    "Secondary education (vs. primary/none) — changed because too few had graduate degrees",
    "Added post hoc; history of anxiety had high prevalence (33.9%) in the sample",
    "P-values reported for all secondary outcomes in Table 2"
  ),
  `SAP language` = c(
    "\"The age cutoff was originally set at 80 years, but this was changed to 75 years due to the low number of participants aged 80 years or older.\" (Results paper)",
    "\"The educational-level category was adjusted from graduate to secondary education.\" (Results paper)",
    "\"An additional unplanned analysis of history of anxiety was included due to its high prevalence.\" (Results paper)",
    "SAP states secondary outcomes are exploratory; results paper reports p-values in Table 2 without explicit multiplicity disclaimer in the table"
  ),
  check.names = FALSE
)

# Print with reasonable formatting
for (i in seq_len(nrow(changes))) {
  cat(sprintf("Change %d: %s\n", i, changes$Feature[i]))
  cat(sprintf("  Planned:   %s\n", changes$`Protocol/SAP specified`[i]))
  cat(sprintf("  Published: %s\n", changes$Published[i]))
  cat(sprintf("  Evidence:  %s\n\n", changes$`SAP language`[i]))
}
```

### 4b. Are these changes problematic?

Not all protocol deviations are equal. Consider each on a spectrum:

**Pre-specified but pragmatically adjusted (low concern):**
The age and education subgroup cutoff changes were both driven by the actual
sample composition — there were simply too few participants in the original
categories to analyze. The authors disclosed these changes explicitly in the
Results section. This is transparent post-hoc adaptation, not hidden manipulation.

**Added post hoc (moderate concern):**
The history of anxiety subgroup was not pre-specified. The authors added it
because anxiety had a high prevalence in their sample and the effect appeared
notable. This is a legitimate scientific observation but it means the reported
subgroup finding has a higher false-positive rate than a pre-specified subgroup.
It should be interpreted as hypothesis-generating.

**Reporting practice inconsistency (moderate concern):**
The SAP described secondary outcomes as exploratory and noted that p-values
for secondary outcomes would not serve as the basis for conclusions. Yet p-values
are reported in Table 2 without an explicit "exploratory only" label on each row.
A reader who does not consult the SAP may treat the secondary p-values as
confirmatory tests. The Methods section does note the exploratory nature, but
the table itself does not.

::: {.callout-important}
**The broader lesson:** Even well-conducted, transparently reported trials in
prestigious journals contain protocol deviations. This is normal — clinical
research requires pragmatic adaptation. The critical-appraisal skill is to
identify and evaluate these deviations, not to condemn the study for having them.
A deviation disclosed in the paper is far less concerning than a deviation you
discover only by reading the protocol.
:::

---

## 5. Sample Size and Power: What Was Planned vs. What Was Observed

The IVAM-ED protocol specified the sample size using three inputs: α = 0.05,
power = 90%, and an effect size of **0.68** (Cohen's d) for the primary outcome.
Assuming 20% dropout, this produced the target of **112 participants** (56 per group).

This section works through that calculation, examines what happens if the true
effect size is smaller than planned, and addresses a common but invalid practice:
*post-hoc power analysis*.

```{r}
#| label: power-setup

# Install pwr if needed: renv::install("pwr")
library(pwr)
```

### 5a. Reproducing the planned sample size

```{r}
#| label: planned-power

# Reproduce the protocol's sample size calculation
# Two-sample t-test, two-sided, α = 0.05, power = 90%, effect size d = 0.68
pwr_planned <- pwr.t.test(
  d    = 0.68,
  sig.level = 0.05,
  power     = 0.90,
  type      = "two.sample",
  alternative = "two.sided"
)
pwr_planned

# Completers needed (before adding dropout allowance)
cat(sprintf("\nCompleters per group needed: %.0f\n", ceiling(pwr_planned$n)))
cat(sprintf("With 20%% dropout: %.0f per group enrolled\n",
    ceiling(pwr_planned$n / 0.80)))
cat(sprintf("Published target: 56 per group (112 total)\n"))
```

::: {.callout-note}
**Interpreting the power calculation:** The trial was powered to detect an effect
of d = 0.68 with 90% probability if that effect truly exists. The 0.68 effect size
was based on a prior study of a comparable psychological intervention. With the
20% dropout adjustment, 56 participants per group were enrolled — consistent with
the published n = 112.
:::

### 5b. Sensitivity analysis: what if the true effect is smaller?

```{r}
#| label: power-sensitivity
#| fig-width: 7
#| fig-height: 4

# Power curve: what is the probability of a significant result across a range
# of true effect sizes, given n = 52 completers per group (observed)?
n_completers <- 52  # observed completers per group

d_seq   <- seq(0.10, 1.0, by = 0.05)
power_seq <- sapply(d_seq, function(d) {
  pwr.t.test(n = n_completers, d = d, sig.level = 0.05,
             type = "two.sample", alternative = "two.sided")$power
})

power_df <- data.frame(d = d_seq, power = power_seq)

ggplot(power_df, aes(x = d, y = power)) +
  geom_line(linewidth = 1, color = "#4472C4") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = 0.90, linetype = "dotted", color = "#C00000") +
  geom_vline(xintercept = 0.68, linetype = "dashed", color = "#538135") +
  annotate("text", x = 0.68, y = 0.15, label = "Planned\nd = 0.68",
           hjust = -0.1, color = "#538135", size = 3.2) +
  annotate("text", x = 0.15, y = 0.81, label = "80% power",
           hjust = 0, color = "grey40", size = 3.2) +
  annotate("text", x = 0.15, y = 0.91, label = "90% power (planned)",
           hjust = 0, color = "#C00000", size = 3.2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title   = "Power to Detect a Significant SRQ-20 Effect (n = 52 completers/group)",
    x       = "True effect size (Cohen's d)",
    y       = "Probability of significant result (power)",
    caption = "Dashed grey = 80% threshold; red dotted = planned 90%; green = protocol effect size"
  ) +
  theme_minimal(base_size = 12)
```

::: {.callout-warning}
**What the power curve tells you:** At the planned effect size (d = 0.68), the
study has approximately 90% power — consistent with the protocol. But if the true
effect is closer to d = 0.40 (a plausible smaller effect for a brief 12-week
behavioral intervention), power drops to roughly 55–60%. More than one in three
real effects of that size would be missed. This is the fragility that modest
effect sizes in small trials can create.

This does not mean the trial was underpowered or that the result is wrong. It means
that the conclusion "the intervention works" should be held with appropriate
uncertainty about magnitude, and that future trials should consider larger samples
or longer follow-up if the effect is expected to be smaller than the planning estimate.
:::

### 5c. The post-hoc power fallacy

After seeing a statistically significant result with an observed mean difference
of approximately −1.2 points and a pooled SD of about 4.5, one might be tempted
to compute the "observed power" — plugging the observed effect size into the power
formula to see whether the study was "adequately powered."

**This is invalid and should never be done.**

```{r}
#| label: posthoc-power-demo

# Observed Cohen's d from our ANCOVA estimate
model_check <- lm(srq20_12 ~ group + srq20_0, data = filter(dat, missing_12wk == 0))
obs_md      <- coef(model_check)["group"]
pooled_sd   <- sd(filter(dat, missing_12wk == 0)$srq20_0)
obs_d       <- obs_md / pooled_sd

cat(sprintf("Observed MD: %.2f\n", obs_md))
cat(sprintf("Pooled SD:   %.2f\n", pooled_sd))
cat(sprintf("Observed d:  %.2f\n", obs_d))

# 'Post-hoc power' using the observed d — this is the fallacy
posthoc_pwr <- pwr.t.test(n = 52, d = abs(obs_d), sig.level = 0.05,
                           type = "two.sample", alternative = "two.sided")
cat(sprintf("\n'Post-hoc power' (invalid): %.1f%%\n", posthoc_pwr$power * 100))
```

::: {.callout-important}
**Why post-hoc power is a fallacy — and what to do instead.**

Post-hoc power takes the observed effect size and asks "how powerful was this study
to detect an effect *this large*?" The answer is mathematically determined by the
p-value: any result with p < 0.05 will have post-hoc power above approximately 50%;
any result with p > 0.05 will have post-hoc power below 50%. The calculation adds
no information.

The circular reasoning: we observed this effect size precisely because the study
was significant. If the study had been non-significant (and the observed effect
smaller), post-hoc power would be lower. Post-hoc power tracks significance —
it does not independently assess whether the study was large enough.

**What to do instead:**
- Report the **prospective power** from the protocol's sample size calculation
  (which is what matters for design evaluation).
- For a non-significant result, report a **confidence interval** and discuss
  whether the upper bound rules out a clinically important effect.
- For future trials, use a **sensitivity power analysis** (as in Section 5b above):
  show what effect sizes the proposed design can and cannot reliably detect.

The American Statistical Association and most major clinical journals explicitly
discourage post-hoc power analysis. The key reference is Hoenig & Heisey (2001),
*The Abuse of Power: The Pervasive Fallacy of Power Calculations for Data Analysis*,
*The American Statistician*, 55(1):19–24.
:::

---

## 6. Practice Exercises

1. **Is the SRQ-20 effect clinically significant?** Using the MCID of 3 points,
   write a two-sentence interpretation of the primary outcome result that conveys
   both what the data show and what they do not show.

2. **Apply Bonferroni to SCI-R.** If the SCI-R (self-care) p-value from your
   analysis is below 0.05, is it also below the Bonferroni-corrected threshold
   of 0.01? What language would you use to describe this result in a paper?

3. **Read the protocol SAP section.** Open
   `source/extracted/matzenbacher-2026-supplement-1.md` and find the statistical
   analysis plan section. Identify the sentence that describes how secondary
   outcome p-values would be reported. Does the published Table 2 follow that
   guidance exactly?

4. **Evaluate the post-hoc anxiety subgroup.** The results paper reports that
   participants with a history of anxiety benefited more (MD −3.72 on SRQ-20)
   than those without (MD −0.81). This subgroup was added post hoc. Without a
   pre-specified hypothesis, what are two reasons to be cautious about this
   finding?

5. **Design the next trial.** Based on the IVAM-ED findings, what would you
   change in the design of a confirmatory follow-up trial? Consider: sample size,
   outcome selection, subgroup pre-specification, follow-up duration.

---------------------------------------------------------------------
---

## Summary

In this final module you:

- Calculated Cohen's d and compared the effect size to an approximate MCID,
  distinguishing statistical significance from clinical significance.
- Identified the single pre-specified primary outcome and labeled the secondary
  outcomes as exploratory.
- Applied a Bonferroni correction and observed how the multiplicity threshold
  changes the interpretation of secondary findings.
- Compared the published analysis to the pre-registered protocol and identified
  four specific deviations, evaluating the concern level of each.

**Completing the sequence:** You have now followed the IVAM-ED trial from data
structure (Module 1) through descriptive statistics (Module 2), visualization
(Module 3), unadjusted estimation (Module 4), ANCOVA (Module 5), missing data
(Module 6), and critical appraisal (Module 7). Each module builds on the previous
one using the same dataset — illustrating how a complete analysis of a randomized
trial unfolds from raw data to published conclusion.

---------------------------------------------------------------------
---

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
