---
title: "Module 8: Capstone — Write a Methods and Results Section"
subtitle: "CRP 241 — IVAM-ED Teaching Case"
author: "Your Name"
date: today
execute:
  echo: true
  warning: false
  message: false
---

## Overview

The seven preceding modules have taught you to handle data, describe samples,
visualize distributions, estimate effects, model with ANCOVA, handle missing data,
and critically appraise findings. This capstone exercise asks you to bring those
skills together in a single task that closely mirrors real research work:

**Write a complete Methods and Results section for the primary outcome of the
IVAM-ED trial, as if you were the corresponding author.**

There is no new code to introduce. The challenge is synthesis — using what you
have already computed to produce coherent, accurate, publication-ready prose.

---

## Background: What a Methods and Results Section Looks Like

In a randomized clinical trial report, the Methods section describes *what was
done* (design, participants, intervention, outcomes, analysis). The Results section
describes *what was found* (flow of participants, baseline characteristics, primary
outcome, secondary outcomes). Together they answer: "What did the investigators
do, and what happened?"

The target length for this exercise is approximately **400–600 words** total
(Methods + Results). Brevity and precision are both required. Every number
should match your analysis output. Every analytical choice should be described
accurately.

---

## Part 1: Data Analysis

Run the complete analysis below before writing. All the numbers you need will
appear in the output.

```{r}
#| label: setup

library(dplyr)
library(gtsummary)
library(mice)
```

```{r}
#| label: load-data

dat        <- readRDS("../data/ivam_working.rds")
completers <- filter(dat, missing_12wk == 0)
```

### Participant flow

```{r}
#| label: participant-flow

cat("Enrolled:         ", nrow(dat), "\n")
cat("  Intervention:   ", sum(dat$group == 1), "\n")
cat("  Usual care:     ", sum(dat$group == 0), "\n")
cat("Missing at 12 wk:", sum(dat$missing_12wk), "\n")
cat("  Reason - death:          ", sum(dat$reason_missing == "death", na.rm=TRUE), "\n")
cat("  Reason - lost to f/u:    ", sum(dat$reason_missing == "lost_to_followup", na.rm=TRUE), "\n")
cat("Completers:       ", nrow(completers), "\n")
cat("  Intervention:   ", sum(completers$group == 1), "\n")
cat("  Usual care:     ", sum(completers$group == 0), "\n")
```

### Baseline characteristics

```{r}
#| label: baseline-table

tbl_summary(
  dat |> select(group_label, age, female, edu_years, dm_duration_y,
                insulin, mmse_0, srq20_0, hba1c_0),
  by = group_label,
  statistic = list(all_continuous() ~ "{mean} ({sd})",
                   all_categorical() ~ "{n} ({p}%)"),
  digits   = list(all_continuous() ~ 1),
  missing  = "no",
  label    = list(age ~ "Age, y", female ~ "Female", edu_years ~ "Education, y",
                  dm_duration_y ~ "Diabetes duration, y", insulin ~ "Insulin use",
                  mmse_0 ~ "MMSE score", srq20_0 ~ "SRQ-20 (baseline)",
                  hba1c_0 ~ "HbA1c, %")
) |> add_overall() |> bold_labels()
```

### Primary outcome: ANCOVA

```{r}
#| label: primary-ancova

model_base <- lm(srq20_12 ~ group + srq20_0, data = completers)
model_full <- lm(srq20_12 ~ group + srq20_0 + age + female +
                   edu_years + income_wages + mmse_0,
                 data = completers)

# Baseline-adjusted
base_coef <- coef(summary(model_base))["group", ]
base_ci   <- confint(model_base)["group", ]

# Fully adjusted
full_coef <- coef(summary(model_full))["group", ]
full_ci   <- confint(model_full)["group", ]

# Adjusted group means at 12 weeks
grand_bl  <- mean(completers$srq20_0)
adj_means <- predict(model_base,
  newdata = data.frame(group = c(0, 1), srq20_0 = c(grand_bl, grand_bl)),
  se.fit = TRUE)

cat("=== Baseline-adjusted ANCOVA ===\n")
cat(sprintf("  Control adj mean (SE):       %.2f (%.2f)\n",
    adj_means$fit[1], adj_means$se.fit[1]))
cat(sprintf("  Intervention adj mean (SE):  %.2f (%.2f)\n",
    adj_means$fit[2], adj_means$se.fit[2]))
cat(sprintf("  MD (int - ctrl):             %.2f\n", base_coef["Estimate"]))
cat(sprintf("  95%% CI:                      (%.2f, %.2f)\n",
    base_ci[1], base_ci[2]))
cat(sprintf("  p-value:                     %.3f\n\n", base_coef["Pr(>|t|)"]))

cat("=== Fully adjusted ANCOVA ===\n")
cat(sprintf("  MD:    %.2f\n", full_coef["Estimate"]))
cat(sprintf("  95%% CI: (%.2f, %.2f)\n", full_ci[1], full_ci[2]))
cat(sprintf("  p-value: %.3f\n", full_coef["Pr(>|t|)"]))
```

### Missing data sensitivity

```{r}
#| label: bocf-sensitivity

dat_bocf <- dat |>
  mutate(srq20_12_bocf = if_else(is.na(srq20_12), srq20_0, srq20_12))
model_bocf <- lm(srq20_12_bocf ~ group + srq20_0, data = dat_bocf)
bocf_coef  <- coef(summary(model_bocf))["group", ]
bocf_ci    <- confint(model_bocf)["group", ]

cat(sprintf("BOCF sensitivity: MD = %.2f (%.2f, %.2f), p = %.3f\n",
    bocf_coef["Estimate"], bocf_ci[1], bocf_ci[2], bocf_coef["Pr(>|t|)"]))
```

---

## Part 2: Write the Methods Section

Using the analysis output above and your knowledge of the study design, write a
Methods section covering the following elements. Your prose should be accurate,
concise, and written in the past tense.

**Required elements:**

1. **Study design:** State the design (randomized clinical trial), setting, and
   enrollment period.
2. **Participants:** State the key eligibility criteria (age ≥ 65, type 2 diabetes)
   and the enrolled sample size.
3. **Randomization:** Describe how participants were assigned (1:1 ratio, computer-
   generated simple random sequence, no stratification or blocking).
4. **Intervention:** Describe what the smart speaker group received and for how long.
   State that the control group received usual care.
5. **Primary outcome:** Name the SRQ-20, state its range and directionality (higher =
   more distress), and state that it was assessed at 12 weeks.
6. **Statistical analysis:** State that the primary analysis was a baseline-adjusted
   ANCOVA with group and baseline SRQ-20 as predictors. State the significance
   threshold (α = 0.05). Describe the BOCF sensitivity analysis.
7. **Missing data:** State the number missing and the assumed mechanism (MAR).

---

## Part 3: Write the Results Section

Using the numbers from your analysis output, write a Results section covering the
following elements. Every number must match your output — not the published paper
(you are reporting the synthetic dataset).

**Required elements:**

1. **Participant flow:** State how many were enrolled, how many completed follow-up,
   and the reasons for dropout.
2. **Baseline characteristics:** Summarize the groups briefly (one or two sentences
   comparing key variables). Do not say the groups were "significantly different"
   on baseline characteristics — state whether any differences are clinically notable.
3. **Primary outcome:** Report the adjusted mean (SE) for each group at 12 weeks,
   the mean difference (95% CI), and the p-value from the baseline-adjusted ANCOVA.
   Then report the fully adjusted model result.
4. **Sensitivity analysis:** State that the BOCF analysis produced a consistent
   finding.
5. **Interpretation sentence:** Write one sentence interpreting what the result
   means clinically, including a comparison to the approximate MCID of 3 points.

---

## Part 4: Peer Review Checklist

After drafting your Methods and Results, apply this checklist before submitting.

```{r}
#| echo: false

checklist <- data.frame(
  Item = c(
    "All numbers match the analysis output (not the published paper)",
    "Design, setting, and enrollment period are stated in Methods",
    "Randomization procedure is described accurately",
    "ANCOVA model covariates are listed",
    "Participant flow numbers are correct and consistent",
    "Baseline comparison uses clinically meaningful language (no p-values)",
    "Adjusted means are reported with SEs, not raw means",
    "Mean difference is reported with 95% CI and p-value",
    "BOCF sensitivity result is reported",
    "Interpretation sentence addresses clinical magnitude (MCID)"
  ),
  `Done?` = rep("☐", 10),
  check.names = FALSE
)

knitr::kable(checklist)
```

---

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
