---
title: "Answer Key"
subtitle: "Selected Exercise Solutions — Modules 1–8"
---

# Answer Key {.unnumbered}

This appendix provides worked solutions for selected practice exercises from each
module. Answers are keyed to the synthetic dataset generated by `R/simulate-ivam-ed.R`
with `set.seed(241)`. **Exact numbers will differ if you use a different seed.**

::: {.callout-note}
These answers reflect one correct approach, not the only approach. Students who
arrive at equivalent results through different code are correct. Instructors
should treat these as reference solutions, not rigid rubrics.
:::

---

## Module 1: Data Setup {.unnumbered}

**Exercise 1 (Recalculate by hand):** Group means for SRQ-20:

```{r}
#| label: m1-ex1

library(dplyr)
dat <- readRDS("../data/ivam_working.rds")

# Base R approach (no packages)
tapply(dat$srq20_0, dat$group_label, mean, na.rm = TRUE) |> round(2)
```

**Exercise 2 (Education):** The simulation targets 24.1% with secondary education
or higher (from the published Results text). Check your synthetic data:

```{r}
#| label: m1-ex2

mean(dat$edu_years >= 8, na.rm = TRUE) * 100  # ~8 years ≈ secondary education
```

The percentage will differ slightly from 24.1% because `edu_years` was simulated
from a continuous normal distribution and the threshold is approximate.

**Exercise 3 (Most imbalanced variable):** Compute SMDs from Module 2 and find the
largest. Typically SRQ-20 shows the largest imbalance (intervention group ~1 SD
higher at baseline due to random assignment with this seed).

**Exercise 4 (SRQ-20 imbalance):** The intervention group's higher baseline SRQ-20
means that unadjusted follow-up means will be closer together than the true treatment
effect — regression to the mean partially offsets the intervention benefit in the
raw comparison. This is exactly the motivation for ANCOVA (Module 5).

---

## Module 2: Descriptive Statistics {.unnumbered}

**Exercise 1 (Recalculate):**

```{r}
#| label: m2-ex1

completers <- filter(dat, missing_12wk == 0)

completers |>
  group_by(group_label) |>
  summarise(mean_srq20_0 = round(mean(srq20_0), 2),
            sd_srq20_0   = round(sd(srq20_0),   2))
```

**Exercise 2 (Education percentage):**

```{r}
#| label: m2-ex2

mean(dat$edu_years >= 8) * 100 |> round(1)
# Published: 24.1% with secondary or higher
```

**Exercise 3 (Most imbalanced SMD):**

```{r}
#| label: m2-ex3

library(gtsummary)

dat |>
  select(group_label, srq20_0, pss_0, sf36_0, hba1c_0, mmse_0) |>
  tbl_summary(by = group_label,
              statistic = list(all_continuous() ~ "{mean} ({sd})"),
              missing = "no") |>
  add_difference(test = list(all_continuous() ~ "smd")) |>
  modify_header(estimate ~ "**SMD**")
```

**Exercise 4 (Why no p-values in Table 1):** In a randomized trial, any baseline
difference arose by chance — by definition, because we randomized. Testing whether
it is "statistically significant" is circular: we already know it is due to chance.
A small p-value for a baseline difference means only that the sample was large
enough to detect a chance imbalance, not that the imbalance is important.

---

## Module 3: Visualization {.unnumbered}

**Exercise 2 (Regression to the mean):** Participants with SRQ-20 > 15 at baseline
tend to have lower scores at follow-up in *both* groups. This is regression to the
mean — extreme values at enrollment move toward the population average over time,
regardless of treatment. It would be wrong to attribute all of the reduction to the
intervention in the smart speaker group; some of it would occur anyway.

**Exercise 3 (PSS trajectory plot):**

```{r}
#| label: m3-ex3
#| fig-width: 8
#| fig-height: 5

library(ggplot2)

pss_long <- completers |>
  select(participant_id, group_label, pss_0, pss_12) |>
  tidyr::pivot_longer(c(pss_0, pss_12), names_to = "timepoint", values_to = "pss") |>
  mutate(timepoint = factor(recode(timepoint, pss_0 = "Baseline", pss_12 = "12 weeks"),
                            levels = c("Baseline", "12 weeks")))

ggplot(pss_long, aes(x = timepoint, y = pss, group = participant_id,
                     color = group_label)) +
  geom_line(alpha = 0.3, linewidth = 0.5) +
  stat_summary(aes(group = group_label), fun = mean, geom = "line",
               linewidth = 1.5) +
  scale_color_manual(values = c("Usual care" = "#ED7D31",
                                "Smart speaker" = "#4472C4")) +
  labs(title = "PSS Trajectories", x = NULL, y = "PSS score", color = NULL) +
  theme_minimal(base_size = 12) + theme(legend.position = "bottom")
```

The visual separation between group means at 12 weeks should look smaller than for
the SRQ-20, consistent with the PSS result not being statistically significant in
Module 4.

---

## Module 4: Estimation {.unnumbered}

**Exercise 1 (Difference in differences):**

```{r}
#| label: m4-ex1

completers |>
  mutate(srq20_change = srq20_12 - srq20_0) |>
  group_by(group_label) |>
  summarise(mean_change = round(mean(srq20_change), 3)) |>
  summarise(diff_in_diff = diff(rev(mean_change)))  # smart - usual

# Compare to unadjusted between-group MD at follow-up
completers |>
  group_by(group_label) |>
  summarise(mean_fu = mean(srq20_12)) |>
  summarise(md_fu = diff(rev(mean_fu)))
```

The two numbers should be identical. Algebra: (µ_int_12 − µ_int_0) − (µ_ctrl_12 −
µ_ctrl_0) = µ_int_12 − µ_ctrl_12 − (µ_int_0 − µ_ctrl_0). If baseline means are
equal, the difference in differences equals the follow-up mean difference.

**Exercise 4 (Cohen's d):**

```{r}
#| label: m4-ex4

t_res     <- t.test(srq20_12 ~ group_label, data = completers, var.equal = FALSE)
pooled_sd <- completers |> summarise(sd = sd(srq20_0)) |> pull(sd)
m_smart   <- t_res$estimate[grep("Smart", names(t_res$estimate))]
m_usual   <- t_res$estimate[grep("Usual",  names(t_res$estimate))]
d         <- (m_smart - m_usual) / pooled_sd

cat(sprintf("Cohen's d = %.2f\n", d))
cat(sprintf("Protocol planned d = 0.68\n"))
cat(sprintf("Interpretation: %s\n",
    ifelse(abs(d) < 0.2, "small",
    ifelse(abs(d) < 0.5, "small-medium", "medium"))))
```

The observed d is smaller than the protocol's planned 0.68 (a common finding when
the true effect is more modest than the planning estimate).

---

## Module 5: ANCOVA {.unnumbered}

**Exercise 1 (Report the group coefficient):**

*"In the baseline-adjusted ANCOVA model, participants in the smart speaker group
had a mean SRQ-20 score [X] points lower (95% CI: [lo], [hi]; p = [p]) at 12 weeks
compared with the usual care group, after adjusting for baseline mental distress."*

Replace the bracketed values with your model output.

**Exercise 3 (Adjusted mean by hand):**

```{r}
#| label: m5-ex3

model_base <- lm(srq20_12 ~ group + srq20_0, data = completers)
grand_mean <- mean(completers$srq20_0)
alpha      <- coef(model_base)["(Intercept)"]
beta_bl    <- coef(model_base)["srq20_0"]

# For group = 0, y0 = grand_mean:
by_hand <- alpha + 0 * coef(model_base)["group"] + beta_bl * grand_mean
by_pred <- predict(model_base,
  newdata = data.frame(group = 0, srq20_0 = grand_mean))

cat(sprintf("By hand:   %.3f\n", by_hand))
cat(sprintf("predict(): %.3f\n", by_pred))
```

**Exercise 4 (Degrees of freedom):** Baseline-adjusted model: 103 − 3 = 100 df.
Fully adjusted model: 103 − 8 = 95 df. Fewer df means slightly wider confidence
intervals and slightly less power — a small cost for the precision gained from
including additional covariates.

---

## Module 6: Missing Data {.unnumbered}

**Exercise 1 (MCAR vs MAR):** Missing participants had higher baseline SRQ-20 scores
than completers. This supports MAR, not MCAR. Under MCAR, missingness would be
unrelated to any variable; here it is related to an observed baseline value. The
MAR assumption is that missingness is related to *observed* variables (baseline
SRQ-20) but not to the unobserved follow-up value — which is plausible but not
verifiable from the data.

**Exercise 4 (Widest CI):** BOCF typically produces the widest CI among the three
approaches because it imputes identical baseline values for missing participants,
inflating both group means toward the population average and reducing the apparent
between-group difference. Multiple imputation should produce a slightly narrower
CI than complete-case analysis when the imputation model is well-specified.

---

## Module 7: Interpretation {.unnumbered}

**Exercise 1 (Clinical significance):**

*"The smart speaker intervention reduced mean SRQ-20 scores by approximately [X]
points compared with usual care (95% CI: [lo], [hi]). Although this reduction was
statistically significant, the point estimate is below the approximate minimum
clinically important difference of 3 points, and the upper bound of the confidence
interval is close to zero, suggesting the clinical benefit may be modest for
individual patients."*

**Exercise 5 (Design the next trial):** A well-reasoned answer would include:
- Larger sample: powered for d ≈ 0.40 rather than 0.68 (requires ~200/group for 80% power)
- Longer follow-up: 24 or 52 weeks to assess durability
- Pre-specify fewer secondary outcomes or apply multiplicity correction
- Include the post-hoc anxiety subgroup as a pre-specified stratification factor
- Multi-site design to improve generalizability beyond a single Brazilian academic center

---

## Module 8: Capstone {.unnumbered}

A complete exemplar Methods and Results paragraph is provided below.
Exact numbers will differ from yours because this uses one seed — compare
structure and language rather than values.

**Methods (exemplar):**

*"The IVAM-ED trial was an open randomized clinical trial conducted at a single
academic medical center. Adults aged 65 years or older with type 2 diabetes were
randomly assigned in a 1:1 ratio using a computer-generated simple random sequence,
without stratification or blocking, to receive a smart speaker (Amazon Echo Dot,
3rd generation) programmed with a behavioral intervention for 12 weeks or to continue
with usual care. The primary outcome was the Self-Reporting Questionnaire-20
(SRQ-20; range 0–20, higher scores indicating greater mental distress) assessed at
12 weeks. The primary analysis was a baseline-adjusted analysis of covariance with
randomized group and baseline SRQ-20 score as covariates. Statistical significance
was set at α = 0.05. A baseline-observation-carried-forward sensitivity analysis
was conducted to assess the impact of missing data (8% of participants)."*

**Results (exemplar, numbers approximate):**

*"Among 112 enrolled participants (56 per group), 103 (92%) completed the 12-week
follow-up; 3 participants died and 6 were lost to follow-up. Baseline demographic
and clinical characteristics were broadly similar between groups. In the baseline-
adjusted analysis, the adjusted mean (SE) SRQ-20 score at 12 weeks was 7.22 (0.43)
in the usual care group and 6.76 (0.45) in the smart speaker group, yielding a mean
difference of −0.47 (95% CI: −2.06 to 1.12; p = .56) [note: your values will
differ]. In the fully adjusted model including age, sex, education, income, and MMSE,
the mean difference was [X] (95% CI: [lo], [hi]; p = [p]). The BOCF sensitivity
analysis produced a consistent result (MD = [X], 95% CI: [lo], [hi]). Although the
direction of effect favored the smart speaker group, the magnitude of the estimated
difference was smaller than the approximate minimum clinically important difference
of 3 points."*

---

## Session Information {.unnumbered}

```{r}
#| label: session-info

sessionInfo()
```
